---
title: "Data Cleaning"
output: html_document
date: "2024-11-07"
---

# Load necessary libraries

```{r}

library(dplyr)

```

# Load dataset

```{r}

newo_data <- read.csv("neworleans_listings.csv")

```

# Select useful variables

- We could use JSON to code amenities variable
- Need to look more into the following columns:
  - amenities
  - minimum_nights 
  - maximum_nights
  - minimum_minimum_nights
  - maximum_minimum_nights
  - minimum_maximum_nights
  - maximum_maximum_nights
  - minimum_nights_avg_ntm
  - maximum_nights_avg_ntm  
  - maybe create a column with average rating as a global rating score to use as y
  - decide which is our y

```{r}

filtered_data <- newo_data %>% 
  dplyr::select(host_since, host_location, host_listings_count, host_response_time,
                host_response_rate, host_acceptance_rate, host_is_superhost,
                host_total_listings_count, host_has_profile_pic, host_identity_verified,
                latitude, longitude, room_type, accommodates, bathrooms, beds, price, 
                #minimum nights, maximum nights, minimum nights avg ntm, maximum nights avg ntm,
                number_of_reviews, review_scores_rating, review_scores_accuracy,
                review_scores_cleanliness, review_scores_checkin, review_scores_communication, 
                review_scores_location, review_scores_value)

```

# Replace empty values or N/A values by NA and count how many missing values per column

```{r}

# Replace "N/A" with NA in the entire dataset
filtered_data <- filtered_data %>%
  mutate(across(c(host_response_time, host_response_rate, host_acceptance_rate), ~ na_if(., "N/A")))

# Replace empty with NA in the entire dataset
filtered_data <- filtered_data %>%
  # Apply `na_if()` to character columns only
  mutate(across(c(host_location, host_is_superhost, host_acceptance_rate), ~ na_if(., ""))) %>%
  # Clean and convert `price` to numeric if it contains characters
  mutate(price = as.numeric(gsub("[$,]", "", as.character(price))))

# Look at $ of NA per column
na_counts <- colSums(is.na(filtered_data))
na_counts

```
# Convert T/F to 1/0 --> true = 1, false = 0

```{r}

filtered_data <- filtered_data %>%
  mutate(across(c(host_is_superhost, host_has_profile_pic, host_identity_verified),
                ~ if_else(as.character(.) == "t", 1, if_else(as.character(.) == "f", 0, NA_real_))))

```

# Code categorical variables with different levels as dummy variables

```{r}

# Room Type
table(filtered_data$room_type)

# Create dummy varaible to code room_type column
dummies_roomtype <- model.matrix(~ room_type - 1, data = filtered_data)

# Convert the matrix to a data frame and add it to the original dataset
filtered_data <- cbind(filtered_data, dummies_roomtype)

# Host Response Type
table(filtered_data$host_response_time)

# Create dummy varaible to code host_response_time column
filtered_data$host_response_time[is.na(filtered_data$host_response_time)] <- "unknown"

# Create dummy variables including the "unknown" category
dummies_responsetime <- model.matrix(~ host_response_time - 1, data = filtered_data)

# Add the dummy variables back to the data
filtered_data <- cbind(filtered_data, dummies_responsetime)

colnames(filtered_data) <- gsub(" ", "_", colnames(filtered_data))
colnames(filtered_data) <- gsub("/", "_", colnames(filtered_data))

filtered_data <- filtered_data %>%
  rename(entire_unit = room_typeEntire_home_apt,
         hotel = room_typeHotel_room,
         private_room = room_typePrivate_room,
         shared_room = room_typeShared_room)

filtered_data <- filtered_data %>%
  rename(resp_time_days = host_response_timea_few_days_or_more,
         resp_time_unknown = host_response_timeunknown,
         resp_time_day = host_response_timewithin_a_day,
         resp_time_hours = host_response_timewithin_a_few_hours,
         resp_time_one_hour = host_response_timewithin_an_hour)

```

# Modify host_since column to host for x years, easier to model

```{r}

filtered_data <- filtered_data %>%
  mutate(
    # Remove "host since " and extract the year part
    host_since_year = gsub("host since ", "", host_since),
    host_since_year = as.Date(host_since_year),            # Convert to Date
    host_years = ifelse(!is.na(host_since_year),            # Check if the date is valid
                        as.numeric(format(Sys.Date(), "%Y")) - as.numeric(format(host_since_year, "%Y")),
                        NA)                               # If not valid, set as NA
  )

```

# Create final dataset

```{r}

final_data <- filtered_data %>%
  select(-host_since, -room_type, -host_since_year, -host_response_time)

# What can we do with this column?
#table(filtered_data$host_location)

```